---
title: "Estimating the effect of number of remaining teeth on social participation among older adults in Japan"
author: "Upul Cooray, Georgios Tsakos, Anja Heilmann, Richard G Watt, Katsunori Kondo, Kenji Takeuchi, Ken Osaka & Jun Aida"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y at %H:%M')`"
output:
  word_document:
    reference_docx: style_doc/ije-style.docx
bibliography: Bib_files/references.bib
csl: CSL_file/CSE-author-date.csl
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)


inline_hook <- function(x) {
  if(is.numeric(x)){
    if(abs(x-round(x)) < .Machine$double.eps){
      # treat as integer
      formatted <- format(x, digits= 0, big.mark= " ")
      
    } else{
      
      formatted <-format(x, digits= 1, nsmall=1)
    }
    
  }else{ 
    formatted <- x
    
    } 
  
  return(formatted)
  
  }



knitr::knit_hooks$set(inline = inline_hook)
  

library(magrittr)
library(tidyverse)
library(targets)
```

## Introduction

The term "social participation" refers to an individual's involvement in
activities that allow them to interact with others in society or the
community [@levasseur2010]. Social participation among older adults is
regarded as an essential component of healthy ageing because it has
numerous positive effects on both individuals and society
[@Golinowska2016]. Previous studies have linked higher levels of social
participation to higher life expectancy, better health-related quality
of life, well-being, and functioning of older adults [@Wanchai2018;
@dahan2008]. Community-level health promotion and prevention activities
such as improving physical activity, smoking and alcohol interventions
could also be facilitated through social engagement [@Saito2019]. On the
other hand, a wide range of determinants, including health-related
factors, have been found to influence older adults' level of social
participation [@Cornwell2009].

Teeth and oral health are important in speaking, smiling, and making
facial expressions, all of which are essential for positive social
interactions. Tooth loss is highly prevalent among older adults as a
result of a life-long accumulation of chronic dental conditions such as
dental caries and periodontal diseases [@Griffin2012]. Previous studies
have consistently linked social and neighbourhood related factors such
as social capital and social participation to oral health related
outcomes among older adults [@Takeuchi2013; @Rouxel2015; @Aida2011].
However, it is equally important to understand the effect of oral health
on participating in social activities. To date, no studies have looked
into the effect of having more natural teeth on social participation in
older adults. Experimental studies to evaluate the potential causal
impact of the remaining number of teeth is not practically feasible.
Additionally, it is complicated by the time varying nature of the number
of teeth and the confounders. We tried to capture the potential causal
effect of number of remaining teeth on social participation using
longitudinal observational data leveraging on *longitudinal modified
treatment policy (LMTP)* approach to specify the causal effect.

LMTP is a recently developed non-parametric alternative that can be used
to define causal effects [@Daz2021]. The literature for causal inference
based on binary exposures is extensive. However, dichotomisation of the
exposure using a arbitrary cut point leads to loss of information on the
exposure and impair the ability to observe any "dose-response" effect on
the outcome. LMTP, on the other hand, allows us to quantify the effect
of a treatment that changes the observed level of exposure in each
individual to a new level [@Daz2021] . In other words, this framework
can be adapted to quantify counterfactual outcomes for questions such
as, "What would have happened to the prevalence of social participation
if everyone in the study population increased or decreased their number
of teeth by a certain amount?", and "What if everyone loss their
teeth?", utilising observational data. Furthermore, the corresponding
statistical estimates for LMTP can be estimated using sophisticated
doubly-robust statistical estimators, such as the targeted minimum
loss-based estimation (TMLE) with Super Learner , which allows for the
use of flexible machine learning predictions avoiding parametric
modelling assumptions [@Laan2018; @schuler2016].

This study presents an analysis that estimates the effect of number of
remaining teeth on social participation among older adults while taking
the time-varying nature of variables into account. LMTP was used to
dynamically shift the observed level of exposure (number of remaining
natural teeth) to a new level in order to investigate its effect on
social participation among functionally independent older adults in
Japan over a 6-year period. A higher number of teeth would, presumably,
have a positive impact on social participation. Hence, we hypothesised
that as the number of teeth increases, social participation improves,
and as the number of teeth decreases, social participation declines
among older adults.

```{r include=FALSE}

base_cov <- c("Age","Sex","L0_inc", "Y0_any","L0_den", "L0_mari","L0_srh")
l1_cov <- c("L1_inc", "L1_den", "L1_mari", "L1_srh")
expo <- c("A0_teeth","A1_teeth")
out <- "Y2"

df <- targets::tar_read(descriptive_data)

df_nomis_expo <- df %>% filter_at(vars(expo),~!is.na(.))
df_nomis_out_expo <- df %>% filter_at(vars(c(expo,out)),~!is.na(.))

df_nomis_base <- df_nomis_out_expo %>%filter_at(vars(base_cov),~!is.na(.))
df_nomis_base_l1 <- df_nomis_out_expo %>%filter_at(vars(base_cov,l1_cov),~!is.na(.))

# Number of eligible participants
n_baseline <- nrow(df)

n_nomis_out_expo <- nrow(df_nomis_out_expo)

missing_expo <- n_baseline- nrow(df_nomis_expo)

missing_out <- n_baseline- nrow(df_nomis_out_expo)- missing_expo

# Number of people with missing baseline cavariates
missing_base_cov<- n_nomis_out_expo - nrow(df_nomis_base)

# Number of people with missing follow-up covariates
missing_l1_cov <-   n_nomis_out_expo - missing_base_cov- nrow(df_nomis_base_l1)

# Checking numbers
# nrow(analytic_data)+ missing_base_cov+ missing_l1_cov==n_baseline

# Get the number censored at c1
c1_censored <- df_nomis_base %>% filter(c1==0) %>% nrow()

#Get the number censored at c2
c2_censored <- df_nomis_base_l1 %>% filter(c1==1 & c2==0) %>% nrow()

followed_l1 <- df_nomis_base %>% filter(c1==1) %>% nrow()

followed_end <- df_nomis_base_l1 %>%  filter(c1==1 & c2==1) %>% nrow()

n_at_2013 <-  n_nomis_out_expo -missing_base_cov

n_at_2016 <- nrow(df_nomis_base_l1)

# censoring ---------


n_died_13 <- df_nomis_base %>% dplyr::filter(C1==5) %>% nrow()

n_ineligible_13 <- df_nomis_base %>% dplyr::filter(C1==3|C1==4) %>% nrow()

n_lost_13 <- df_nomis_base %>% dplyr::filter(C1==6) %>% nrow()

# Check
# n_died_13+ n_ineligible_13+ n_lost_13==c1_censored

n_died_16 <- df_nomis_base_l1 %>% dplyr::filter(c1==1 & C2==5) %>% nrow()

n_ineligible_16 <- df_nomis_base_l1 %>% dplyr::filter(c1==1 & (C2==3|C2==4)) %>% nrow()

n_lost_16 <- df_nomis_base_l1 %>% dplyr::filter(c1==1 & C2==6) %>% nrow()

```

## Methods

### Data

Data from the Japan Gerontological Evaluation Study (JAGES) was used in
this study [@Kondo2018]. JAGES is an on-going nationwide cohort study
for functionally independent older adults in Japan aged 65 years or
over. For this analysis, data from the 2010 survey as the baseline and
two subsequent follow-up surveys (2013 and 2016) were used. A total of
`r n_baseline` functionally independent individuals were identified at
the baseline survey. Out of them `r n_2016` individuals had responded to
all three waves of JAGES (i.e. 2010, 2013, and 2016). During the 6 years
of follow-up `r n_died_13+n_died_16` had died,
`r n_ineligible_13+ n_ineligible_16` became ineligible as they became
functionally dependent, and `r n_lost_13+ n_lost_16` were lost to
follow-up due to other reasons. The flow of study participants within
three waves is shown in Figure 1. A comparison of baseline
characteristics by participants follow-up status (i.e. died/ became
ineligible/ lost to follow-up/ or remained) is reported in Supplementary
Table 1.

### Outcome variable

Social participation in 2016 was the outcome in this study. JAGES
recorded the frequency of social participation ("nearly every day,"
"twice or thrice a week," "once a week," "once or twice a month," "a few
times/year," "never") for various social activities. We assessed the
frequency of participation in any of the following activities: hobby
groups, sports clubs, senior citizens' clubs, residence groups, or
volunteer groups. Participation in any of the aforementioned activities
once a month or more frequently (vs. less frequently or never) was
defined as indicative of social participation (1 = participation, 0 =
non-participation) [@shiba2021].

### Exposure

The number of remaining natural teeth at the time of the surveys in 2010
and 2013 was used as a time-varying exposure in the analysis. The
self-reported number of teeth was recorded using the response to the
question, "How many natural teeth do you currently have?" (Instructions:
capped/crowned teeth should be counted as "natural teeth"). The
responses of participants were recorded in four categories (i.e., 20
teeth/ 10-19 teeth/ 1-9 teeth/ no teeth).

### Covariates

Because the number of teeth was evaluated as a time-varying exposure in
the study, both time-invariant and time-variant covariates were taken
into account. As time-invariant covariates, age (range 65--99 years),
sex (male/female), and social participation in 2010 (outcome at the
baseline) were adjusted for. Equalised annual household income (million
yen), self-rated health (very good/ good/ fair/ poor), denture use (yes/
no), and marital status (married/ single, widowed or divorced) were used
as time-variant covariates (measured in 2010 and 2013).

### Statistical analysis

The hypothesised temporal connections between study variables are shown
in the directed acyclic graph (Figure 2). A descriptive analysis was
performed to identify the characteristics of participants stratified by
the outcome (social participation in 2016). Then, to specify the impact
of number of teeth on counterfactual outcomes, the observed level of
number of teeth of each individual was shifted to several new levels
mimicking multiple hypothetical interventions. Following hypothetical
scenarios were created by shifting the observed exposure to detect any
dose-response associated with the outcome.

1\. "all participants having ≥20 teeth in 2010 and in 2013," (i.e. ideal
counterfactual scenario where all the participants having a minimal
functional dentition [@gotfredsen2007])

2\. "all participants having 10-19 teeth in 2010 and in 2013,"

3\. "all participants having 1-9 teeth in 2010 and in 2013,"

4\. "all participants being edentate in 2010 and in 2013," (i.e. worse
counterfactual scenario where all the older adults were edentate)

5\. "observed level of number of teeth category" (i.e. originally
observed data in 2010 and in 2013).

To estimate the social participation with the shifted (and the observed)
exposure, we used TMLE [@diaz; @Schuler2016]. In TMLE, the probability
of the exposure conditional on covariates (exposure model), and the
conditional probability of the outcome given exposure and covariates
(outcome model) were estimated to obtain an unbiased estimation of the
counterfactual outcomes [@Laan2006; @Schuler2016]. If either the
exposure model or the outcome model was consistently estimated, unbiased
estimates could be obtained (hence doubly-robust) [@Laan2012]. To
increase the likelihood of robust specification of exposure and outcome
models, Super Learner, an ensemble method that uses weighted
combinations of multiple machine learning algorithms was used
[@Laan2007; @Rose2019; @Schomaker2019]. Within Super Learner,
generalized linear models (glm), extreme gradient boosting models, and
neural nets were used as candidate algorithms [@Venables2002;
@Chen2016]. Then, the TMLE estimate for the 'ideal' counterfactual
scenario (i.e. everyone having a functional dentition at each time
point) was used as a reference to calculate causal odds ratios (OR) and
95% confidence intervals (95% CI) for other respective scenarios.
Additionally, ORs and 95% CIs were calculated using the TMLE estimate
for the observed level of number of teeth as the reference. To check the
robustness of TMLE estimates, an analysis without using Super Learner
(using only glm) was also conducted. All estimates were appropriately
controlled for aforementioned time-variant and time-invariant
covariates. E values for the estimates were calculated to report the
impact of unmeasured confounders. Furthermore, information on censoring
was included in models to reduce bias due to attrition of the study
population.

For the imputation of missing data in covariates, we used random forest
multivariate imputation by chained equations (MICE). In imputing complex
epidemiologic data, random forest MICE has been shown to produce less
biased parameter estimates and better confidence interval convergence
compared to parametric MICE [@Shah2014]. Analyses were performed using
five imputed datasets and the results were pooled using Rubin's rules.
Random forest MICE was implemented using `mice` R package [@Buuren2011].
The distribution of missingness among covariates in relation to outcome
and exposure are reported in Supplementary Figure 1 and 2 . A
supplementary complete case analysis was conducted to check the
robustness if results with imputed data (Appendix xx). The `lmtp` R
package was used to compute TMLE estimates with Super Learner for each
scenario [@lmtpR]. Main R functions used to generate our results are
provided in Appendix xx. All the other codes that used for data
preparation and supplementary analyses can be found at
`https://github.com/upulcooray/social-participation`. All the analyses
were conducted in R studio using `R version 4.1.2` for x86_64,
linux-gnu.

## Results

```{r}
prev_non_part <- df$Y2 %>% table() %>% .[[1]]


```

Baseline characteristics of the sample stratified by the outcome
variable are presented in Table 1. In the 2016 follow-up,
`r prev_non_part` people reported a social participation frequency of
less than at least once a month. Baseline characteristics associated
with less frequent social participation in 2016 were older age, lower
income, poor self-reported health, being edentate, and lower frequency
of baseline social participation.

Table 2 reports the causal odds ratios calculated by contrasting TMLE
estimates related to observed and four hypothetical scenarios. Results
indicate that having a relatively lower number of teeth negatively
affected social participation after the six-year follow-up. After
adjusting for age, sex, baseline social participation and time-varying
confounders such as annual household income, self-rated health, dental
prosthesis use, and marital status, consistent exposure to
edentulousness would have lowered the mean social participation by 9%
(OR= 0.91 & 95%CI= 0.84,0.99). Similarly, the odds of social
participation were lower if the dental status of everyone in the sample
population deteriorated by one step (see hypothetical intervention No.5
in the methods section) in terms of number of teeth category (OR= 0.91 &
95%CI= 0.86,0.96). On the other hand, continuously being in the 20 or
more teeth category (having a minimal functional dentition) increased
the odds of social participation by 9% (OR= 1.09 & 95%CI= 1.07,1.11).

Figure caption: Odds ratio for social participation under different
hypothetical interventions (no intervention as the reference): OR,95%
CI, p-value Results related to the effect of dental prosthesis use on
social participation for the different number of teeth categories are
shown in Figure
xx.Figure caption: Odds ratio for social participation for different
number of teeth categories under hypothetical denture status levels
(observed denture status as the reference): OR,95% CI, p-value The
overall trend of the results indicated that dental prostheses had a
positive effect on social participation across all number of teeth
categories. However, the findings revealed that a significant beneficial
effect of denture use on social participation could only be found among
the 10-19 and 1-9 teeth categories (OR= 1.16 & 95%CI= 1.05,1.27; OR=
1.19 & 95%CI= 1.01,1.38, respectively). Social participation among the
edentate group and those with more than 20 teeth did not appear to be
affected by denture status.

## Discussion 

Using an analytic approach that enabled assessing the effect of
different levels of an exposure over time while adjusting for
time-variant covariates, we estimated the effect of dental status on
social participation. A doubly-robust estimator (TMLE) in combination
with machine learning based ensemble (SuperLearner) was used to obtain
the estimates. To the best of our knowledge, this is the first study to
estimate the effect of number of remaining natural teeth on social
participation. Our analysis revealed that having a relatively higher
number of teeth during the follow-up period improved the social
participation among older adults in our study population. Being edentate
and experiencing consistent tooth loss during the follow-up period
reduced social participation. Furthermore, our findings suggest that
wearing dentures would improve social participation among older adults
with 1-9 teeth or 10-19 teeth. Dentures did not appear to provide a
significant benefit to edentate people in terms of improving social
participation. Our findings are consistent with prior studies related to
the link between oral health and social participation $$\@$$. However,
previous studies were based on cross-sectional data and treated oral
health as the dependent variable. Our analysis gave insight into how
number of teeth would affect social participation using longitudinal
data. Given the evidence that social participation improves the health
and well-being of older adults, any mechanism that leads to increased
levels of social participation should be promoted and encouraged $$\@$$.
In that context, our findings emphasise the importance of older adults
retaining a higher number of teeth, not only for obvious benefits on
oral functions such as mastication and speech, but also to have better
social relationships and thus reap the benefits associated with social
participation. The analysis, which was intended to investigate if the
dental prostheses could improve the social participation, identified
that the benefit of dentures was limited to those who had 1-9 teeth or
10-19 teeth. These two groups are unique as these groups represent
relatively unstable status of the dentition compared to individuals in
\>=20 teeth category or edentate state. It is plausible that these two
groups are more susceptible to the challenges posed by impaired
oral functions as a result of tooth loss. Edentate people, on the other
hand, transition into their edentate state over a long time period, so
their baseline level of social participation may already be established,
making them less likely to be affected by dental prostheses. Overall,
these findings suggest that preventing tooth loss in order to maintain a
minimum functional dentition in older adults would be more effective
than providing dental prostheses in terms of improving social
participation. The use of a doubly-robust TMLE estimator to estimate the
effects of multiple hypothetical interventions that are otherwise
unmeasurable, can be considered the study's main strength. Furthermore,
our estimates were based on longitudinal data while allowing for the
time varying nature of the exposure and confounders. However, several
limitations should be noted. Variables used in this study were based on
self-reported measures, which are prone to measurement and
classification errors. However, previous studies conducted in Japan have
shown the validity of self-reported number of teeth [@Matsui2016;
@Ueno2018]. Causal inference for time-varying exposure necessitates no
unmeasured confounding assumption at each time point (conditional
exchangeability assumption). Despite the fact that we adjusted estimates
for multiple time-varying and time-invariant confounders, as well as
baseline levels of social participation, the possibility of unmeasured
confounding cannot be ruled out. In our analyses we used panel data with
participants who took part in all three waves of the JAGES. Therefore,
we had a large attrition of our sample population (n= 52053 at baseline
to n= 22451 at 2016 follow-up). We examined the baseline characteristics
associated with attrition (supplementary table xx) and showed that it
was associated with a lower number of teeth at baseline. However, a low
number of teeth negatively affected social participation, thus, any
selection bias caused due to attrition would have led to an
underestimation of the effects of tooth loss on social participation. In
this study, we only looked at the effect of number of teeth on the
frequency of social participation. However, the quality of social
interactions is just as important in accruing the health benefits of
social participation $$\@$$. It is reasonable to assume that the quality
of social interactions is more affected by teeth and oral health than
the frequency of participation. Therefore, further research is warranted
to investigate this assumption. Moreover, we were not able to identify
the locations of missing teeth in our data. Missing front teeth would
have had a greater impact on facial aesthetics and speech, whereas
missing posterior teeth (molars and premolars) would have mainly
impacted masticatory functions. Thus, the location of missing teeth
would have affected differently on social participation. These effects
should be further investigated in future studies. Conclusions
