---
title: "Effect of number of teeth on social participation among older adults in Japan: using modified treatment policy approach"

output:
  word_document:
    reference_docx: style_doc/ije-style.docx
bibliography: Bib_files/references.bib
csl: CSL_file/international-journal-of-epidemiology-2.csl
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)


inline_hook <- function(x) {
  if(is.numeric(x)){
    if(abs(x-round(x)) < .Machine$double.eps & x>=10000){
      # treat as integer
      formatted <- format(x, digits= 0, big.mark= " ",
                          big.interval = 3L)
    } else{
      
      formatted <-format(x, digits= 1, nsmall=1)
    }
  }else{ 
    formatted <- x
    } 
  
  return(formatted)
  }

knitr::knit_hooks$set(inline = inline_hook)
  

library(magrittr)
library(tidyverse)
library(targets)
```

```{r include=FALSE}

df <- targets::tar_read(descriptive_data)

  n_baseline <- nrow(df)

  n_2013 <-  df %>% filter(c1==1) %>% nrow()

  n_2016 <-  df %>% filter(c1==1 & c2==1) %>% nrow()

  # censoring ---------

  n_died_13 <- df %>% dplyr::filter(C1==5) %>% nrow()

  n_ineligible_13 <- df %>% dplyr::filter(C1==3|C1==4) %>% nrow()

  n_lost_13 <- df %>% dplyr::filter(C1==6) %>% nrow()

  n_c1 <- n_died_13+n_ineligible_13+ n_lost_13

  # Check
  # n_died_13+ n_ineligible_13+ n_lost_13+ n_2013==c1_censored

  n_died_16 <- df %>% dplyr::filter(c1==1 & C2==5) %>% nrow()

  n_ineligible_16 <- df %>% dplyr::filter(c1==1 & (C2==3|C2==4)) %>% nrow()

  n_lost_16 <- df %>% dplyr::filter(c1==1 & C2==6) %>% nrow()

  n_c2 <-  n_died_16+n_ineligible_16+ n_lost_16
  
  prev_non_part <- df$Y2 %>% table() %>% .[[1]]
  percen_prev <- prev_non_part/sum(!is.na(df$Y2))*100

```

```{r}


# res<- tar_read(pooled_estimates)
res<- tar_read(results_sl)

res<- res %>% 
  mutate(or_ci= 
           str_replace(
             `OR [95% CI]`, 
             "^(\\d+\\.\\d+)\\s\\[(.*)]",
             "OR=\\1, 95%CI=\\2")) %>% 
  select(contrast, or_ci)

obs_vs_s1 <- res$or_ci[[1]]
obs_vs_s2 <- res$or_ci[[2]]
obs_vs_s3 <- res$or_ci[[3]]
obs_vs_s4 <- res$or_ci[[4]]
obs_vs_s5 <- res$or_ci[[5]]



```

\*Upul Cooray^1^, Georgios Tsakos^2^, Anja Heilmann^2^, Richard G
Watt^2^, Kenji Takeuchi^1^, Katsunori Kondo^4,5^, Ken Osaka^1^ & Jun
Aida^3^

^1^Department of International and Community Oral Health, Tohoku
University Graduate School of Dentistry, Sendai, Japan

^2^Department of Epidemiology and Public Health, University College
London, London, United Kingdom

^3^Department of Oral Health Promotion, Graduate School of Medical and
Dental Sciences, Tokyo Medical and Dental University, Tokyo, Japan 

^4^Center for Preventive Medical Sciences, Chiba University, Chiba,
Japan

^5^Center for Gerontology and Social Science, National Center for
Geriatrics and Gerontology, Obu, Japan

### **\*Corresponding Author:**

Upul Cooray

Department of International and Community Oral Health, Graduate School
of Dentistry, Tohoku University, Sendai, Japan 

Address: 4-1, Seiryo-machi, Aoba-ku, Sendai, Miyagi, 980-8575, Japan

E-mail: [upul.cooray.15\@ucl.ac.uk](mailto:upul.cooray.15@ucl.ac.ukp),
Tel: +81-2-2717-7639

Abstract word count: 252

Total word count: 3063

Number of tables: 2

Number of figures: 3

Number of reference: 34

\
**Keywords**: Casual inference, Modified treatment policy, Targeted
minimum loss-based estimation, Social participation, Older adults

\

\pagebreak

## Abstract (word limit 250)

### Background

Participating in social activities and interacting with others in the
community has numerous positive effects on older adults' health and
quality of life. We aimed to estimate the causal effect of number of
teeth on social participation among older adults in Japan.

### Methods

Using longitudinal data (baseline=2010, follow-ups=2013 and 2016) from
`r n_2016` participants of the Japan Gerontological Evaluation Study, we
employed a longitudinal modified treatment policy approach to determine
the causal effect of number of teeth on social participation.
Corresponding statistical parameters were estimated using targeted
minimum loss-based estimation (TMLE). Number of teeth (edentate, 1-9
teeth, 10-19 teeth, $\geq$ 20 teeth) was treated as a time-varying
exposure and the estimates were adjusted for time-varying (income,
self-rated health, denture use, marital status) and time-invariant (age,
sex, baseline social participation) covariates.

### Results

After six years of follow-up, `r percen_prev`% reported less frequent
social participation (less than once a month). Causal odds ratios,
calculated by contrasting counterfactual TMLE estimates at different
levels of the exposure, showed a clear dose-response effect on social
participation. When the exposure was shifted from being edentate to
having $\geq$ 20 teeth, the likelihood of social participation increased
by 30% (`r eden_vs_mfd`). A shift to all participants being edentate
reduced the social participation by 15% (`r obs_vs_eden`).

### Conclusions

This study provides causal evidence that having a higher number of teeth
and maintaining a functional dentition positively affects social
participation among Japanese older adults, while being edentate or
having relatively fewer teeth negatively affects social participation.

\pagebreak

## Introduction

The term "social participation" refers to an individual's involvement in
activities that allow them to interact with others in their community or
society in general [@levasseur2010]. Social participation among older
adults is an essential component of healthy ageing because it has
numerous positive effects on both individuals and society
[@Golinowska2016]. Previous studies have linked higher levels of social
participation to higher life expectancy, better health-related quality
of life, well-being, and functioning of older adults [@Wanchai2018;
@dahan2008; @holt-lunstad2010]. Community-level health promotion and
prevention activities such as physical activity, smoking and alcohol
interventions, could also be facilitated through social engagement
[@Saito2019]. On the other hand, a wide range of determinants, including
health-related factors, influence older adults' level of social
participation [@Cornwell2009; @zhang2020].

Teeth and oral health are important in different aspects of daily life,
such as eating, speaking, smiling, and making facial expressions, all of
which are essential for positive social interactions. Tooth loss is
highly prevalent among older adults due primarily to a life-long
accumulation of chronic dental conditions such as dental caries and
periodontal diseases [@Griffin2012]. Previous studies have consistently
linked social and neighbourhood related factors such as social capital
and social participation to oral health related outcomes among older
adults [@Takeuchi2013; @Rouxel2015; @Aida2011]. much less is known about
the effect of oral health on participation in social activities.
Experimental studies to evaluate the potential causal effect of the
remaining number of teeth on social participation are not practically
feasible, further complicated by the time varying nature of the exposure
(number of teeth) and confounders. Longitudinal modified treatment
policy *(LMTP)* approach can be adapted to overcome some of these
limitations for causal inference with observational data [@ikeda2022].

LMTP is a recently developed non-parametric alternative that can be used
to define causal effects [@Daz2021]. The literature for causal inference
based on binary exposures is extensive [@höfler2005]. However,
dichotomisation of the exposure using a arbitrary cut-off point leads to
loss of information on the exposure and hinders the ability to observe
any "dose-response" effect on the outcome. LMTP, on the other hand,
allows us to quantify the effect of a treatment that changes the
observed level of exposure in each individual to a new level [@Daz2021].
In other words, this framework can be adapted to quantify counterfactual
outcomes for questions such as, "What would have happened to the
prevalence of social participation if everyone in the study population
increased or decreased their number of teeth by a certain amount?", and
"What if everyone in the study population lost their teeth?".
Furthermore, the corresponding statistical parameters for LMTP can be
estimated using sophisticated doubly-robust statistical estimators, such
as the targeted minimum loss-based estimation (TMLE) with Super Learner,
which allows for the use of flexible machine learning predictions
avoiding parametric modelling assumptions [@Laan2018; @schuler2016].

This study estimates the effect of the number of remaining teeth on
social participation among older adults while taking the time-varying
nature of variables into account. LMTP was used to dynamically shift the
observed level of exposure (number of remaining natural teeth) to new
levels in order to investigate its effect on social participation among
functionally independent older adults in Japan over a 6-year period. We
hypothesised that as the number of teeth increases, social participation
improves, and as the number of teeth decreases, social participation
declines among older adults.

## Methods

### Data

Data from the Japan Gerontological Evaluation Study (JAGES) was used in
this study [@kondo2018]. JAGES is an on-going nationwide cohort study
for functionally independent older adults in Japan aged 65 years or
over. For this analysis, data from the 2010 survey as the baseline and
two subsequent follow-up surveys (2013 and 2016) were used. From a total
of `r n_baseline` functionally independent participants in the baseline
survey, `r n_2016` individuals responded to all three waves (i.e., 2010,
2013, and 2016). During the 6 years of follow-up `r n_died_13+n_died_16`
died, `r n_ineligible_13+ n_ineligible_16` became ineligible as they
became functionally dependent, and `r n_lost_13+ n_lost_16` were lost to
follow-up due to other reasons (study flow chart in Figure 1). A
comparison of baseline characteristics by participants' follow-up status
(i.e., died/ became ineligible/ lost to follow-up/ or remained) is
reported in Supplementary Table S1.

### Outcome variable

Social participation in 2016 was the outcome in this study. JAGES
recorded the frequency of social participation ("nearly every day",
"twice or thrice a week", "once a week", "once or twice a month", "a few
times/year", "never") for various social activities. We assessed the
frequency of participation in any of the following activities: hobby
groups, sports clubs, senior citizens' clubs, residence groups, or
volunteer groups. Participation in any of the aforementioned activities
once a month or more frequently (vs. less frequently or never) was
defined as indicative of social participation (1=participation,
0=non-participation) [@shiba2021].

### Exposure

The number of remaining natural teeth at the time of the surveys in 2010
and 2013 was used as a time-varying exposure in the analysis. The
self-reported number of teeth was recorded using the response to the
question, "How many natural teeth do you currently have?" (Instructions:
capped/crowned teeth should be counted as "natural teeth"). The
responses of participants were recorded in four categories (i.e., 20
teeth/ 10-19 teeth/ 1-9 teeth/ no teeth).

### Covariates

Because the number of teeth was evaluated as a time-varying exposure in
this study, both time-invariant and time-variant covariates were taken
into account. As time-invariant covariates, age (range 65--99 years),
sex (male/female), and social participation in 2010 (outcome at the
baseline) were adjusted for. Equalised annual household income (million
yen), self-rated health (very good/ good/ fair/ poor), denture use (yes/
no), and marital status (married/ single, widowed or divorced) were used
as time-varying covariates (measured in 2010 and 2013).

### Statistical analysis

The hypothesised temporal associations between study variables are shown
in the directed acyclic graph (Figure 2). A descriptive analysis was
performed to identify the characteristics of participants stratified by
the outcome (social participation in 2016). Then, to specify the impact
of number of teeth on counterfactual outcomes, the observed level of
number of teeth of each individual was shifted to several new levels
mimicking multiple hypothetical interventions. Following hypothetical
scenarios were created by shifting the observed exposure to detect any
dose-response associated with the outcome.

1\. "all participants having ≥20 teeth in 2010 and in 2013," (i.e.,
ideal counterfactual scenario where all the participants having a
minimal functional dentition [@gotfredsen2007])

2\. "all participants having 10-19 teeth in 2010 and in 2013,"

3\. "all participants having 1-9 teeth in 2010 and in 2013,"

4\. "all participants being edentate in 2010 and in 2013," (i.e., worse
counterfactual scenario where all the older adults were edentate)

5\. "observed level of number of teeth category" (i.e., originally
observed level in 2010 and in 2013).

To estimate the social participation with the shifted (and the observed)
exposure, we used TMLE [@Daz2021; @schuler2016]. In TMLE, the
probability of the exposure conditional on covariates (exposure model),
and the conditional probability of the outcome given exposure and
covariates (outcome model) were estimated to obtain an unbiased
estimation of the counterfactual outcomes [@van2006; @schuler2016]. If
either the exposure model or the outcome model was consistently
estimated, unbiased estimates could be obtained (hence doubly-robust)
[@van2012]. To increase the likelihood of robust specification of
exposure and outcome models, Super Learner, an ensemble method that uses
weighted combinations of multiple machine learning algorithms was used
[@van2007; @rose2019; @Schomaker2019]. Within Super Learner, generalized
linear models (*glm*), extreme gradient boosting models, and neural nets
were used as candidate algorithms [@nnet; @xgb]. Then, the TMLE estimate
for the 'ideal' counterfactual scenario (i.e., everyone having a
functional dentition at each time point) was used as a reference to
calculate causal odds ratios (OR) and 95% confidence intervals (95% CI)
for other respective scenarios. Additionally, ORs and 95% CIs were
calculated using the TMLE estimate for the observed level of number of
teeth as the reference. To check the robustness of TMLE estimates, an
analysis without using Super Learner (using *glm*) was also conducted.
All estimates were appropriately controlled for aforementioned
time-variant and time-invariant covariates. E values for the estimates
were calculated to report the impact of unmeasured confounders.
Furthermore, information on censoring was included in models to reduce
bias due to attrition of the study population.

For the imputation of missing data in covariates, we used random forest
multivariate imputation by chained equations (MICE). In imputing complex
epidemiologic data, random forest MICE has been shown to produce less
biased parameter estimates and better confidence interval convergence
compared to parametric MICE [@shah2014]. Analyses were performed using
five imputed datasets and the results were pooled using Rubin's rules.
MICE was implemented using `mice` R package [@buuren2011]. The
distribution of missingness among covariates in relation to outcome and
exposure are reported in Supplementary Figure S1 and S2 . A
supplementary complete case analysis was conducted to check the
robustness of results with imputed data. The `lmtp` R package was used
to compute TMLE estimates with Super Learner for each scenario [@lmtpR].
Main R codes used to generate our results are provided in Supplementary
material. All the other codes that used in analyses can be found at
`https://github.com/upulcooray/social-participation`. All the analyses
were conducted in R studio using `R version 4.1.2` for x86_64,
linux-gnu.

## Results

Baseline characteristics of participants stratified by the outcome
variable are presented in Table 1. In the 2016 follow-up,
`r prev_non_part` (`r percen_prev`%) people reported a social
participation frequency of less than at least once a month. Baseline
characteristics associated with less frequent social participation in
2016 were older age, lower income, poor self-reported health, being
edentate, and lower frequency of baseline social participation.

Table 2 provides the causal odds ratios calculated by comparing TMLE
estimates for social participation at various levels of exposure (number
of teeth categories). After a six-year follow-up, the results showed
that having a relatively lower number of teeth had a negative effect on
social participation and having a relatively higher number of teeth had
a positive effect. When the TMLE estimate for the observed level of
exposure was used as the reference, a change from the observed number of
teeth category to an edentate state was negatively associated with a 15%
(`r obs_vs_eden`) reduction in the likelihood of social participation. A
shift from the observed number of teeth category to having 20 or more
teeth had a positive impact on social participation with an
`r obs_vs_mfd` . When the counterfactual outcome under the edentate
state was employed as the reference, a clear positive dose-response
effect of having more natural teeth on social participation was observed
(Figure 3). The highest improvement of the likelihood of social
participation (30%) was observed when the number of teeth category was
shifted from edentate to having 20 or more teeth (`r eden_vs_mfd`). Even
a minor change in number of teeth status, such as going from edentate to
having one to nine teeth, was associated with a 10% (`r eden_vs_1_9`)
increase in the likelihood of frequent social participation. Age, sex,
baseline social participation, and a set of time-varying confounders
(i.e., annual household income, self-rated health, dental prosthesis
use, and marital status) were adjusted for in all reported estimates.
Furthermore, the reported estimates were obtained by allowing exposure
to vary during the follow-up period and accounting for participant
censoring.

Figure 3-a and 3-b shows the clear dose-response in point estimates as
well as the robustness of the results regardless of the use of Super
Learner. However, estimates with Super Learner appear to be slightly
more conservative than without it. Therefore, the TMLE estimates
calculated with Super Learner were used in Table 2 for reporting. The
findings of the supplementary complete case analysis were consistent
with the findings of the main analysis ( Supplementary Table S2 and
Supplementary Figure S3).

## Discussion

To estimate the impact of the number of remaining teeth on social
participation among older adults in Japan, we used an analytic approach
that allows us to estimate the effects of different levels of exposure
over time while controlling for time-variant covariates. The estimates
were obtained using a doubly-robust estimator (TMLE) in combination with
a machine learning-based ensemble (Super Learner). To the best of our
knowledge, this is the first study to estimate the effect of number of
remaining natural teeth on social participation. Our findings show that
having a relatively higher number of teeth during the follow-up period
had a positive effect on social participation among older adults. In
addition, a decrease in the observed number of teeth had a negative
impact on social participation. These findings support our hypothesis
and consistent with previous related research. Previous studies,
however, were based on cross-sectional data and used the number of teeth
as the outcome variable [@Takeuchi2013; @Aida2011]. Using longitudinal
data and causal inference, this study added evidence related to the
importance of maintaining an adequate number of teeth for frequent
social participation among older adults. Given the consistent evidence
that social participation improves older adults' health and well-being,
mechanisms that leads to increased levels of social participation should
be promoted and encouraged. In this context, our findings emphasise the
importance of older adults retaining a greater number of teeth, not only
for obvious benefits on oral functions such as mastication and speech,
but also to have better social relationships and thus reap the benefits
associated with social participation.

The mechanism that explains our findings is straightforward and
intuitive. Teeth play an important role in social interactions such as
smiling, speaking, eating, and maintaining facial aesthetics
[@steele2004]. As a result, tooth loss would naturally lead to a
reluctance to engage in social activities. A recent cross sectional
study by Koyama et al. [@koyama2021] examined the association between
the number of teeth and social isolation among older adults using data
from Japan and England. They found that having fewer teeth was
significantly associated with being socially isolated in both countries.
Although Koyama et al. investigated a different outcome (i.e., social
isolation), the mechanism between the number of teeth and social
isolation may be similar to that of current study. We believe that our
study's estimates, which were based on longitudinal data and a robust
causal inference methodology, add compelling evidence to the link
between oral health and social participation/interactions.

We employed novel LMTP approach to define causal effects without needing
to dichotomise the number of teeth variable (categorical exposure
variable). The LMPT approach naturally satisfied the positivity
assumption [@petersen2010] (i.e., all people had a non-zero probability
of obtaining a specific exposure level) in defining causal effects since
it shifted each individual's exposure level based on their observed
number of teeth level at each time point. Furthermore, by using doubly
robust TMLE to estimate the corresponding statistical parameters, we
were able to minimise parametric modeling assumptions [@rose2019;
@Daz2021]. Traditionally, causal estimates could only be obtained by
contrasting conterfactual outcomes at the exposure's extremes (i.e.
"What if everyone exposed versus everyone not exposed" - due to forced
binary exposures). In this study, for example, a traditional method
would have only allowed us to estimate the difference between the
counterfactual outcomes of being edentate versus having teeth, or having
20 or more teeth versus having less than 20 teeth. However, using the
LMTP/TMLE to estimate counterfactual outcomes across the entire spectrum
of exposure, we were able to detect a gradual increase in social
participation at each level of improvement in the number of teeth
(Figure 3-a). Additionally, TMLE estimate under the observed level of
number of teeth could be contrasted against counterfactual outcomes
under hypothetical exposure levels to estimate the expected change in
mean population outcome (Figure 3-b).

Our analysis has several limitations, some of which may cause estimates
to be biased. First, the variables in this study were self-reported,
which are prone to measurement and classification errors. Previous
studies in Japan, however, have showed the validity of the self-reported
number of teeth measure [@matsui2016; @ueno2012]. Second, causal
inference for time-varying exposure necessitates no unmeasured
confounding assumption at each time point (conditional exchangeability
assumption)[@hernan2006]. Therefore, despite adjusting for multiple
time-varying and time-invariant confounders, as well as baseline levels
of social participation, the possibility of unmeasured confounding
cannot be ruled out (E-values for all point estimates are reported in
Table 2 to reflect the potential effect of unmeasured confounding)
[@vanderweele2017]. Third, we used panel data with older adult
participants who took part in all three waves of the JAGES. Therefore,
we had a large attrition of our sample population within six years(n= 52
053 at baseline to n= `r n_2016` at 2016 follow-up). To minimise the
bias due to this attrition, we included censored participants'
information in our analysis to obtain estimates controlled for
censoring. Additionally, we examined the baseline characteristics
associated with censoring (supplementary table 1), and found that it was
associated with a lower number of teeth at baseline. According to the
study results, having fewer teeth had a negative impact on social
participation. As a result, any selection bias caused by attrition would
have resulted in an underestimation of the effect of number of teeth on
social participation.

Future research is needed to look into some of the missing elements in
our study. We were unable to assess the quality of social interactions
in this study, which is just as important in obtaining the health
benefits associated with social participation. Furthermore, we were
unable to locate the locations of missing teeth in our data. Missing
anterior teeth would have had a greater impact on facial aesthetics and
speech, whereas missing posterior teeth (molars and premolars) would
have had a greater impact on masticatory functions. As a result, the
location of missing teeth would have had a different impact on social
participation.

Despite these limitations, our findings provides robust evidence that
having a higher number of teeth positively affects social participation
among Japanese older adults, while being edentate or having a relatively
lower number of teeth negatively affects their social participation.
This emphasises the importance of incorporating tooth loss prevention
into interventions aimed at increasing social participation among older
adults. Our findings also indicated that increasing the number of teeth
improved social participation, which is comparable to the potential
impact of effective dental prosthetic treatments to restore missing
teeth.

## Conclusions

Having a higher number of teeth positively affects social participation
among Japanese older adults, while being edentate or having a relatively
lower number of teeth negatively affects their social participation.

## References

## 
